---
title: "Incidence prediction"
author: "Maximilian Knoll"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview
This vignette demonstrates the use of the incAnalysis package for the comparative evaluation of model performance for incidence prediction. However, the following approach can be used not only for incidence data but for any kind of rates. 

# Requirements and installation
## INLA 
The package relies on the INLA framework for BAPC models, and INLA might also be used for GLM and GAM fitting. It can be obtained from http://www.r-inla.org/download.

```{r, eval=F}
install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
```

## BAPC
The BAPC package (Riebler2017) is available on R-Forge (https://r-forge.r-project.org/projects/bapc/)

```{r, eval=F}
install.packages("BAPC", repos="http://R-Forge.R-project.org")
```

## incAnalysis 

This package can be easily installed via devtools from github.
```{r eval=F}
if (!any(grepl("devtools", installed.packages()))) {
    install.packages("devtools")
}
devtools::install_github("mknoll/incAnalysis")
```



# Included data
The following cancer registry incidence data, split by gender, are included in the package for selected entities as counts (see below):

- SEER9 data, 1973-2014
- NORDCAN data, all Nordic countries (retreived: October 2018)
- Saarland data (retrieved: October 2018)

Entities:

- Glioblastoma, Lung, Pancreas, Colorectal, Prostate, Breast, Kidney, Melanoma (SEER9)
- All available entities for NORDCAN/Saarland data

Corresponding population data (counts), split by gender, is included as well:

- SEER9: all race population data dived by 1000
- NORDCAN data, all Nordic countries
- Saarland data



# Analysis of GBM incidence data
The analysis workflow will be demonstrated using Glioblastoma multiforme (GBM) incidences from the SEER9 dataset for males aged >= 20. 
Corresponding analyses for NORDCAN and Saarland data are added as non-executed chunks.

```{r}
dataset <- "SEER9" #NORDCAN, Saarland
gender <- "MALE" #FEMALE
entity <- "Glioblastoma" 
ageLow <- 20 # exclude data for ages below 20
```


## Population data 

Population data is named after an equal scheme, thus loading different datasets works by adjusting the dataset variable.

```{r, fig.width=8, fig.height=5}
#filename 
fn <- paste("population_", dataset, "_", gender, ".csv", sep="")
population <- read.csv(
		       system.file("extdata", fn, 
				   package="incAnalysis"), 
		       check.names=F, row.names=1)

#multiply by factor 1000 to get actual counts (only SEER9 data!)
if (dataset == "SEER9") {
    population <- population*1000
}

#visualize population as heatmap
require(pheatmap)
pheatmap(population, cluster_rows=F, cluster_cols=F)

# age/ age categories
rownames(population)

#adjust population matrix rownames
if (dataset == "Saarland") {
    rownames(population) <- do.call(rbind, strsplit(trimws(rownames(population)), " "))[,1]
    rownames(population)[1] <- 0
    population <- data.frame(t(population), check.names=F)
} else if (dataset == "NORDCAN") {
    rownames(population) <- do.call(rbind, strsplit(rownames(population), "-"))[,1]
    population <- data.frame(t(population), check.names=F)
}

population[1:4,1:10]
```

##  Incidence data SEER9
SEER9 data were preprocessed manually and only selected entities are included.

```{r}
#filename 
fn <- paste(dataset, "_", entity ,"_", gender, ".csv", sep="")
incidence <- data.matrix(read.csv(system.file("extdata", fn, package="incAnalysis"), check.names=F, row.names=1))

## additionally included entities
fl <- list.files(system.file("extdata", "/", package="incAnalysis"))
fl[which(grepl("SEER9", fl))]
```

## Incidence data (NORDCAN)

NORDCAN data was acquired via webscraping on 2018-08-01 (all available entities).
```{r, eval=F}
dataset <- "NORDCAN"
load(system.file("extdata", paste(dataset, "_incidence_", gender,sep=""), package="incAnalysis")) #coll
names(coll) <-  unlist(lapply(coll, function(x) x$meta$Name))

#Entities
names(coll)

incidence <- data.matrix(t(coll[which(names(coll) == "Brain, central nervous system")][[1]]$data))
colnames(incidence) <- colnames(population)
```

## Incidence data (Saarland)

Saarland data was acquired via webscraping on 2018-08-01 (all available entities).
```{r, eval=F}
dataset <- "Saarland"
load(system.file("extdata", paste(dataset, "_incidence_", gender,sep=""), package="incAnalysis")) #all
nm <- lapply(all, function(x) x$map$Tumor)
names(all) <- unlist(nm)

#Entities 
names(all)

incidence <- data.matrix(t(all[which(names(all) == "191 Gehirn")][[1]]$data[,]))

## differentially groups in population and incidence -> remove 85+ (incidence) and 85/90 (population)
population <- population[,1:17]
incidence <- incidence[,1:17]
colnames(incidence) <- colnames(population)
```

## Visualize incidence and correct NAs
```{r, fig.width=8, fig.height=5}
# NA correspond to 0 values
incidence[which(is.na(incidence))] <- 0
incidence <- data.frame(incidence, check.names=F)

incidence[1:4,1:10]

#visualize incidence as heatmap
pheatmap(incidence, cluster_rows=F, cluster_cols=F)
```

## Remove childhood cancers
For further analyses, only cancers in persons ages >= ageLow are analyzed.
```{r}
# get ages from colnames 
ages <- as.numeric(gsub("+", "", as.character(colnames(incidence))))
w <- which(ages >= ageLow)

incidence <- incidence[,w]
population <- population[,w]
```

## Create incAnalysis object
```{r}
require(incAnalysis)
obj <- new("incClass", 
	   population=population,
	   cases=incidence,
	   nPred=10, #prediction period: 10 years
	   nFit=15, #fitting period
	   predLast=0, #only retrieve data for last prediction year
	   gridFactor=1, #width of age and year bins. 1 for SEER data, 5 for NORDCAN and Saarland
	   entity=entity)
```

## Forward Projection of Last observations
```{r}
obj <- runFwProj(obj, method="wilson") 
```

## GLMs
Fitted using INLA. 
```{r}
# linear model with age and period effects
frm <- as.formula(Y~offset(log(N))+ PERIOD + AGE)
# use neg-binomial distribution
obj <- runGLM(obj, frm=frm, family="nbinomial", text="age,period", method="inla")
```

## GAMs, unviariate smoother
```{r}
### RW2
frm <- as.formula(Y~offset(log(N))+PERIOD+f(AGE, model="rw2"))
obj <- runGLM(obj, frm=frm, family="nbinomial", text="rw2", method="inla")

### Bspline
require(splines)
frm <- as.formula(Y~offset(log(N))+PERIOD+bs(AGE))
obj <- runGLM(obj, frm=frm, family="nbinomial", text="splineAge", method="inla")
```


## GAMs, multivariate smooths
```{r}
# classical approach (tensor spline)
frm <- as.formula(Y~offset(log(N))+ te(PERIOD, AGE))
obj <- runGAM(obj, frm=frm, family="nb", text="te", method="ml")


### INLA
# marginal bases for age and period
ag <- splines2::mSpline(obj@dataLongPred$AGE)
yr <- splines2::mSpline(obj@dataLongPred$AGE)
  
# tensor product smooth model matrix
Z <- mgcv::tensor.prod.model.matrix(list(ag, yr))
  
# model formula
# SEQNO: unique id for each observation/prediction
frm <- as.formula(Y~offset(log(N)) + AGE + PERIOD
       + f(SEQNO, model="z", Z=Z,
       prior="loggamma", param=c(1, 0.005)))
obj <- runGAM(obj, frm=frm, family="nbinomial", text="multivarte smoother", method="inla")
```

## BAPC
```{r}
obj <- runBAPC(obj)
```

## PIT Histograms
For INLA based analyses
```{r, eval=F}
pitHist(obj)
```

## Calcuate bias, precision, coverage
```{r}
obj <- evaluate(obj)
#to plot predictions, set predLast=1 when instantiating the object
```

## Extract metrics
```{r}
df <- metrics(obj)
df[1:4,]
```

## Plot data
```{r, fig.width=7, fig.height=5}
require(ggplot2)
##retain only future prediction
df$PERIOD_F <-as.numeric(as.character(df$PERIOD))
df <- df[which(df$PERIOD_F > max(df$PERIOD_F)-obj@nPred),]

# Coverage  
ggplot(df, aes(x=PERIOD, y=CVG*100, group=paste(TYPE, TEXT), color=paste(TYPE, TEXT))) + geom_line() + geom_point()+ labs(color="model", y="coverage [%]") + ggpubr::rotate_x_text() + scale_color_brewer(palette="Set1")

# Precision
ggplot(df, aes(x=PERIOD, y=log(SD), group=paste(TYPE, TEXT), color=paste(TYPE, TEXT))) + geom_line() + geom_point() + labs(color="model", y="precision") + ggpubr::rotate_x_text()+ scale_color_brewer(palette="Set1")

# Bias
ggplot(df, aes(x=PERIOD, y=BIAS*100, group=paste(TYPE, TEXT), color=paste(TYPE, TEXT))) + geom_line() + geom_point() + labs(color="model", y="bias [%]") + ggpubr::rotate_x_text()+ scale_color_brewer(palette="Set1")
```
